Upwelling variability across the Benguela upwelling systems

### 2019-10-07 Write_up/Paper1_writeup.Rmd
update

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@Mead; @schlegel]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarred; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel]. Changes in the intensity and duration of upwelling events have proven to cause large changes in ecosystem functioning [@wang].

@@

The Cape peninsula is influenced by south-east trade winds during summer and westerlies during winter, thus a pronounced seasonal cycle is evident within patterns of wind movement across the coastline. During spring, summer, and autumn seasons, the anticyclonic high-pressure cell periodically ridges eastwards south of the continent (RWS: Citation). During this process, south easterly winds predominate, with peak frequencies occurring during spring and summer months resulting in more upwelling events during summer and winter months (RWS: Citation). During the winter seasons, low pressure cells are driven from the west to the east as a result of the atmospheric circumpolar waves and so north-westerly winds dominate during winter. South easterly winds dominate during summer but are still present for short durations during winter [@andrews]. The dynamics of upwelling within the nearshore (<400m from the coastline) remains poorly understood, with Very few studies addressing the changes in the intensity of cold events at varying distances from the coast within the Benguela region.

Cold temperatures are important in describing species population distribution limits and are proven to set limits in the north or southwards higher latitudes (Firth et al., 2011). Upwelling metrics are often calculated far offshore because measurements are typically taken from large ships that cannot move close to the shore [@andrews]. There has not yet been any understanding of how the upwelling process varies at local versus broader scales. Within this study it is hypothesized that the upwelling metric may vary spatially across different distances, with more intense upwelling events occurring nearest to the coast. Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tend to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters [@andrews; @Shannon]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.

@@

Visualisations

@@

>>>>

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents [@Mead; @schlegel]. The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region [@jarred; @hutchings]. The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast [@luning1990; @lutjeharms2000; @hutchings; @schlegel].

@@

The Cape peninsula is influenced by south-east trade winds during summer and westerlies during winter, thus a pronounced seasonal cycle is evident within patterns of wind movement across the coastline. During spring, summer, and autumn seasons, the anticyclonic high-pressure cell periodically ridges eastwards south of the continent (RWS: Citation). During this process, south easterly winds predominate, with peak frequencies occurring during spring and summer months resulting in more upwelling events during summer and winter months (RWS: Citation). During the winter seasons, low pressure cells are driven from the west to the east as a result of the atmospheric circumpolar waves and so north-westerly winds dominate during winter. South easterly winds dominate during summer but are still present for short durations during winter [@andrews]. Within a range of kilometres off the shore, wind conditions vary due to the effects of mountain funnelling within the region (RWS: Citation) and as a result conditions are variable in the nearshore and offshore.

The dynamics of upwelling within the nearshore (<400m from the coastline) remains poorly understood, with Very few studies addressing the changes in the intensity of cold events at varying distances from the coast within the Benguela region. Cold temperatures are important in describing species population distribution limits and are proven to set limits in the north or southwards higher latitudes (Firth et al., 2011). Upwelling metrics are often calculated far offshore because measurements are typically taken from large ships that cannot move close to the shore [@andrews]. There has not yet been any understanding of how the upwelling process varies at local versus broader scales. Within this study it is hypothesized that the upwelling metric may vary spatially across different distances, with more intense upwelling events occurring nearest to the coast.

Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tend to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters [@andrews; @Shannon]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.

@@

@@

# Plotting images

####TO do
- Plots
